FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-22 manuscript.tex
Minor edits by Felix

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8368}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix D. Schönbrodt5},
  pdflang={en-EN},
  pdfkeywords={meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix D. Schönbrodt\textsuperscript{5}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# cca0d45c.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D. Schönbrodt: Conceptualization, Writing -- review \& editing.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany}

@@

Scholars have been raising concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
Two main concerns are that, first, social scientific theories lack precision compared to theories in the physical sciences (Szollosi \& Donkin, 2021).
and clarity
In other words, social scientific theories lack \emph{formalization},
which means that they do not make very accurate predictions,
and are thus hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a philosophical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

\begin{figure}
\centering
\includegraphics{empirical_cycle.pdf}
\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
\end{figure}

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

@@

\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:tabfair}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endhead
F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory (meta)data has a global, unique and persistent identifier & Rephrased\\
F2 & Data are described with rich metadata & Theory is described with rich metadata & Rephrased\\
F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & Rephrased and extended\\
F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & Rephrased, needs work\\
A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & Rephrased\\
A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Remain the same\\
A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Remain the same, but less relevant\\
A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Rephrased, but less relevant\\
I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & Rephrased\\
I2S.1 & - &  & \\
I2S.2 & - &  & \\
I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Extended\\
I4S & - &  & \\
R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory (meta)data are released with a clear and accessible license & Rephrased\\
R1.2 & (Meta)data are associated with detailed provenance & Theory (meta)data are associated with detailed provenance & Rephrased\\
R1.3 & (Meta)data meet domain-relevant community standards & Theory (meta)data and documentation meet domain-relevant community standards & Rephrased\\
\bottomrule
\end{longtable}

@@

There are different definitions of theory,
and many of those definitions are consistent with the FAIR theory principles.
This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data {[}bogen/woodward{]}.
Meehl (1990) provides guidance as to what kinds of ``statements'' such a theory might contain:
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values Guest (2024).

Some have defined a model as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
While there may be philosophical differences,
these differences are largely irrelevant for the purpose of making theories more FAIR.
When organizing knowledge, we think it is helpful to view theories and models derived from them as existing along a continuum of specificity,
from broad to narrow, from general to specific,
where a theory has a relatively broader scope and may contain one or more models as specific instances.
For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.
From this theory, we could derive a more specific \emph{statistical model} by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Or, we could derive an even more specific \emph{generative/computational model}, which is completely parametrized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Note that broadness and narrowness are relative terms,
and one person's theory may be another person's model.
This definition also implies that FAIR models are a special case of FAIR theory.

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) (Taylor, 2022).
The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes that would otherwise get away with their vague formulation in verbal theories.
For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented (Lewandowsky \& Farrell, 2010).
More precise theories makes them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down.
The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory (for example, see the empirical cycle in Figure \ref{fig:figec}),
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
it is not FAIR.
FAIR theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).

@@

@@

Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
It further increases the impact and reuse potential of theories across disciplines,
either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
or through analogical modeling.
In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
Findability also enables meta-research on theories,
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
In a similar way, it would become possible to compare all theories of a specific phenomenon,
or to study structural properties of theories.

The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata (F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
or catalog of psychological phenomena {[}REF Noah Denny{]}.
Metadata should also include identifiers for all the versions of the theory it describes (F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
The Kuhnian notion of ``normal science'', conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding (Kuhn, 2009).
From a pragmatic perspective, these considerations indicate that,
when striving to make theory accessible,
it is important to be as explicit as possible (e.g., about assumptions and ontological definitions),
while acknowledging that accessibility exists on a spectrum,
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, embedding within shared contexts, and explication of assumptions, will advance its Accessibility.

A third impediment arises when theories have a ``dependency on the author'' (DOA).
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the ``ask Leon'' phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on ``Great Man Theorizing'' (Guest, 2024) because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target (Szollosi \& Donkin, 2021).
The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
This demonstrates that original authors are not the ultimate authority on their theories.
DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.

@@

Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as previously recommended (Van Lissa et al., 2021).
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some axiomatic system.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps (Gray, 2017).
While a theory represented as a bitmap image is not very interoperable,
the same image represented in the DOT language for representing graphs does meet this ideal.

@@

Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
we decided to use it as an example for this tutorial.
Note that, while the empirical cycle is not explicitly referred to as a ``theory'' by De Groot, he derives it from ``a theory of thinking'' by Selz (REF).
We can thus consider it a meta-theory of the process of theory construction.
The empirical cycle is described on page 28 of De Groot and Spiekerman (1969):

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8200}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, Felix D. Schönbrodt\textsuperscript{5}, \& Noah N.N. van Dongen\textsuperscript{6}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# 09ba7dc6.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D. Schönbrodt: Conceptualization, Writing -- review \& editing; Noah N.N. van Dongen: Writing -- review \& editing.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{6} University of Amsterdam, the Netherlands}

@@

Scholars have been raising concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One main concern is that social scientific theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
When theories do not make precise predictions,
they are hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a philosophical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \ref{fig:figecs}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

@@

\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:tabfair}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endhead
F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory (meta)data has a global, unique and persistent identifier & Rephrased\\
F2 & Data are described with rich metadata & Theory is described with rich metadata & Rephrased\\
F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & Rephrased and extended\\
F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & Rephrased, needs work\\
A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & Rephrased\\
A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Remain the same\\
A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Remain the same, but less relevant\\
A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Rephrased, but less relevant\\
I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
I2 & (Meta)data use vocabularies that follow FAIR principles & Theory and its associated metadata\ \ use vocabularies that follow FAIR principles & Rephrased\\
I2S.1 & - &  & \\
I2S.2 & - &  & \\
I3 & (Meta)data include qualified references to other (meta)data & Theory and its associated metadata includes qualified references to other theories and their metadata, including previous versions of the theory & Extended\\
I4S & - &  & \\
R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory and its associated metadata are released with a clear and accessible license & Rephrased\\
R1.2 & (Meta)data are associated with detailed provenance & Theory and its associated metadata are associated with detailed provenance & Rephrased\\
R1.3 & (Meta)data meet domain-relevant community standards & Theory and its associated metadata and documentation meet domain-relevant community standards & Rephrased\\
\bottomrule
\end{longtable}

@@

\subsection{What is Theory?}\label{what-is-theory}

Definitions of theory abound and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our paper is not aligned with one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021)
The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's hierarchy of ever-more specific ``statements'' a theory might contain (1990):
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values Guest (2024).

The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term ``model'' is not uniquely defined within the literature;
it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
-

The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g.~commonly used analogies) and practices (e.g.~experimental designs) can be an important part of scientific theories.

Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
the method is not contained to any one particular definition.
It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind of information is useful.
As a particular FAIR theory evolves, the nature of information tracked will likely change.
For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are defined and causally connected.
From this theory, more specific \emph{statistical/mathematical models} could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., TODO), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety; Taylor, 2022).
While the latter psychological factors might be narratively described as vague directional causal statements,
the theory may also encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it involves expressing ideas as precise statements.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the precise implementations of decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

To some extent, FAIRness and formalization are orthogonal.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the precision with which they are communicated.
The FAIR principles pertain to theories' documentation, archival and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).

@@

\subsubsection{Semantic Versioning}\label{semantic-versioning}

Aside from technical solutions, version control is a social process as well.
On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
In computer science, these challenges are navigated by assigning version numbers.
Specifically, \emph{semantic versioning} comprises a simple set of rules for assigning version numbers to information artifacts.
Whereas version control tracks changes,
semantic versioning communicates what those changes mean to users of the theory.
We propose the following adaptation of semantic versioning for theories:

MAJOR version when you make backwards incompatible changes, i.e.,
      the theory now contains empirical statements that are at odds
      with a previous version
MINOR version when you expand the set of empirical statements in a
      backward compatible manner (i.e., the previous version is
      subsumed within the new version)
PATCH version when you make backward compatible bug fixes,
      cosmetic changes, fix spelling errors, or add clarifications
\end{verbatim}

Semantic versioning guides the social process of theory development, communicating how much a theory is changing over tiem.

@@

Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
It further increases the impact and reuse potential of theories across disciplines,
either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
or through analogical modeling.
In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
Findability also enables meta-research on theories,
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
In a similar way, it would become much easier to explicitly compare different theories of a specific phenomenon,
or to study structural properties of theories.

The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
First, this requires assigning a globally unique and persistent identifier (such as a DOI) to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata (F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
or catalog of psychological phenomena {[}REF Noah Denny{]}.
Metadata should also include identifiers for all the versions of the theory it describes (F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
The Kuhnian notion of ``normal science'', conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding (Kuhn, 2009).
From a pragmatic perspective, these considerations indicate that,
when striving to make theory accessible,
it is important to be as explicit as possible (e.g., about assumptions and ontological definitions),
while acknowledging that accessibility exists on a spectrum,
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
will advance its Accessibility (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).

A third impediment arises when theories have a ``dependency on the author'' (DOA).
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the ``ask Leon'' phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on ``Great Man Theorizing'' (Guest, 2024) because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target (Szollosi \& Donkin, 2021).
The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
This demonstrates that original authors are not the ultimate authority on their theories.
DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible in terms of both availability and interpretability.

@@

Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as previously recommended (Van Lissa et al., 2021).
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps (Gray, 2017).
While a theory represented as a bitmap image is not very interoperable,
the same image represented in the DOT language (\textbf{DOTLanguage?}) for representing graphs does meet this ideal.

@@

We will use the empirical cycle as a running example for this tutorial.
The resulting FAIR version of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
The empirical cycle is described on page 28 of De Groot and Spiekerman (1969),
and consists of a series of natural language statements:
